---
title: "Institutional Framing and Sentiment in Cultural News Articles: Evidence from Hespress"
format: 
  html:
    self_contained: true
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
author: Ouissal Akioui
---

# Introduction

This paper examines whether and how institutional framing is associated with sentiment in cultural news coverage in Morocco. It does so using the Moroccan News Articles Dataset (MNAD), a large-scale corpus of Arabic news articles collected from major Moroccan online news outlets. The MNAD corpus was developed to address the scarcity of high-quality, large-scale resources for Arabic text analysis and contains over 418,000 articles distributed across 19 thematic categories. Each article includes a title, a body of text, and a category label, making the dataset particularly well suited for text-as-data approaches that compare discursive patterns across sources while maintaining a consistent unit of analysis at the article level.

The MNAD corpus aggregates content from four Moroccan online news outlets—Akhbarona, Hespress, Hibapress, and Le360—which operate within the same national media environment but differ in editorial practices, communicative conventions, ownership structures, and degrees of institutional proximity. These differences make the dataset a relevant setting for examining how institutional positioning may be reflected in media discourse.

Given the thematic diversity of the dataset, this study deliberately restricts the analysis to articles classified under the *Culture* category. This decision is methodological rather than substantive. News domains such as politics or economics are subject to distinct rhetorical constraints and emotional registers that would confound comparisons of tone. By focusing on cultural coverage, the analysis holds topic relatively constant and reduces variation driven by subject matter, allowing sentiment differences to be more plausibly attributed to editorial and discursive choices rather than to the nature of the news itself.

Within this context, the central research question guiding the analysis is the following: **Is institutional framing associated with systematically more neutral sentiment in cultural news articles?** To address this question, the study adopts a text-as-data framework in which sentiment analysis is used as a measurement tool rather than as an explanatory model. Sentiment scores are computed at the article level, transforming textual content into continuous numerical variables that capture the overall emotional tone of both titles and article bodies. These scores are then used as dependent variables in statistical models designed to assess associations between institutional framing and sentiment, while accounting for basic textual characteristics such as article length.

Importantly, the analysis does not claim to directly measure “state discourse” or authorial intent. Instead, sentiment tone is treated as a proxy for broader discursive patterns commonly associated with institutional communication, such as emotional restraint, neutrality, or formalization. This approach allows for a cautious examination of how institutional positioning may be reflected in media tone without conflating sentiment with ideology or causality.

Based on this framework, the study advances the following hypothesis:

**H1 — Institutional Tone Hypothesis**

Cultural news articles exhibiting stronger institutional framing are expected to display more neutral sentiment than articles with weaker or no institutional framing. The rationale is that institutionally proximate outlets tend to favor emotionally restrained discourse, even outside explicitly political domains.

# Descriptive Analyses & Data Tidying

```{r}
rm(list=ls()) #This line of code cleans the environment

# Load libraries (or packages):
library(tidyverse) 
library(dplyr)

#Load the dataset, we're limiting ourselves to the hespress one which contains the most articles out of the 4:
hespress <- read_csv("Hespress.ma.csv", show_col_types = FALSE)

View(hespress)
```

```{r}
library(dplyr)
library(stringr)

# Create a new dataset containing only articles from the "Culture" category
hespress_culture <- hespress %>%
  mutate(Category = str_trim(Category)) %>%  # Remove any leading or trailing spaces in the Category column
  filter(Category == "Culture")

View(hespress_culture)

```

## Tokenization and stopwords

To analyze the language used in Hespress culture articles, we first break each article down into individual words, a process called tokenization (each word becomes its own unit of analysis). Later on, this allows us to count how often each word appears across the corpus and identify the most common terms.

We tokenize both article titles and article bodies separately, as titles tend to use more condensed, headline-style language while bodies contain the fuller journalistic prose.

```{r}
# Load required packages
# tidyr: data reshaping tools used by tidytext
# tidytext: text tokenization and text mining tools
library(dplyr)
library(tidyr)
library(tidytext)

# Add a unique document identifier for each article. This is necessary so that tokens can be traced back to their original article
hespress_culture <- hespress_culture %>%
  mutate(doc_id = row_number())

# Tokenize the Body text of each article
# Each row of text is split into individual words (tokens)

tokens_body <- hespress_culture %>%
  select(doc_id, Body) %>%  ## Keep only the document ID and article body
  unnest_tokens(         ## Convert text into tokens (one word per row)
    output   = word,
    input    = Body,
    token    = "words",
    to_lower = FALSE 
  )

#same process here
tokens_title <- hespress_culture %>%
  select(doc_id, Title) %>%
  unnest_tokens(
    output   = word,
    input    = Title,
    token    = "words",
    to_lower = FALSE
  )

```

Not all words carry meaning. Common function words like "and", "the", or "in" (called stopwords) appear constantly but tell us nothing about the content. We remove them (the equivalent in Arabic) using a standard Arabic stopword list, and add a few extra words manually (since the list dictionary is not exhaustive) that appeared frequently in our data but were equally uninformative.

```{r}
library(stopwords)
library(dplyr)

# Load a pre-existing Arabic stopword dictionary
# The "misc" source contains common Arabic function words
arabic_sw <- stopwords(language = "ar", source = "misc")

# Add custom stopwords specific to the dataset
manual_sw <- c(
  "عبد",
  "عبر",
  "علي",
  "الي",
  "وذلك",
  "المائة",
  "ذاته",
  "فيما",
  "اخري",
  "اذ",
  "حتى",
  "إلى",
  "بـ",
  "أن",
  "إلا",
  "يذكر",
  "داخل"
)

# Combine built-in and manually defined stopwords
# unique() ensures that duplicates are removed
all_sw <- unique(c(arabic_sw, manual_sw))

# Remove stopwords and very short tokens
# No re-tokenization is needed because the text was already tokenized earlier
tokens_title_clean <- tokens_title %>%
  filter(
    !word %in% all_sw,
    str_length(word) >= 2
  )

# Inspect most frequent words in titles
library(dplyr)
library(knitr)

tokens_title_clean %>%
  count(word, sort = TRUE) %>%
  slice_head(n = 20) %>%
  kable(
    col.names = c("Word", "Frequency"),
    caption = "Top 20 most frequent words in article titles"
  )

```

You’re looking at the 20 most frequent words in article titles in the *culture* corpus. The dominant terms are: National identifiers, cultural domains, institutional / symbolic references. This frequency distribution can indicatej that cultural coverage is strongly anchored in national identity and institutional cultural production, rather than emotive or creative framing.

## Bigrams

Single words can miss context. "Culture" alone tells us less than "Ministry of Culture." Bigrams are pairs of consecutive words, and analyzing them lets us see which concepts tend to appear together in the articles.

```{r}
library(tidytext)
library(dplyr)
library(stringr)

# Create bigrams (two-word combinations) from article bodies
bigrams <- hespress_culture %>%
  select(doc_id, Body) %>% # Keep only the document ID and article body text
  unnest_tokens(             # Tokenize the text into bigrams 
    output = bigram,         
    input  = Body,
    token  = "ngrams",
    n      = 2
  ) %>%
  
  ## Split the bigram into two separate columns
  # This allows us to filter each word individually
  separate(bigram, into = c("word1", "word2"), sep = " ") %>%
  
  # Remove bigrams containing stopwords or extremely short tokens
  # This step removes grammatical words that carry little meaning
  filter(
    !word1 %in% all_sw,
    !word2 %in% all_sw,
    str_length(word1) > 1,
    str_length(word2) > 1
  ) %>%
  ## Recombine the filtered words into a clean bigram
  unite(bigram, word1, word2, sep = " ")


```

```{r}
# Identify the most frequent bigrams in the corpus
bigram_freq <- bigrams %>%
  
  # Count how many times each bigram appears in the dataset
  # sort = TRUE orders the results from most frequent to least frequent
  count(bigram, sort = TRUE) %>%
  
  # Keep the top 20 most frequent bigrams
  slice_head(n = 20)

knitr::kable(
  bigram_freq,
  col.names = c("Bigram", "Frequency"),
  caption = "Top 20 most frequent bigrams in Hespress Culture articles"
)
```

The most frequent bigrams in *Hespress* culture articles are overwhelmingly institutional, national, and formal. Such as **“Ministry of Culture”** (906), **“Mohammed VI”** (845), and **“Minister of Culture”** (513), as well as formal journalistic constructions like **“a statement to Hespress”** (753).

Cultural coverage is thus framed primarily around official actors, state institutions, and nationally symbolic figures and events.

Let's visualise this.

```{r}
library(ggplot2)
# Visualize the most frequent bigrams in the corpus

bigram_freq %>%
  ggplot(aes(x = reorder(bigram, n), y = n, fill = bigram)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Top 20 bigrams in Hespress culture articles",
    x = NULL,
    y = "Frequency"
  ) +
  theme_minimal()


```

This code prepares the data for **bigram network analysis** rather than simple phrase frequency analysis. While the previous code counted how often each two-word expression appears as a single phrase (e.g., *“وزارة الثقافة”*), this version separates each bigram into its two component words (*word1* and *word2*) and counts how often they occur together. By structuring the data this way, the bigrams are treated as relationships between words rather than fixed expressions. Each pair therefore represents a potential edge in a network, where words are nodes and their co-occurrences form connections.

```{r}
bigram_edges <- hespress_culture %>%
  unnest_tokens(bigram, Body, token = "ngrams", n = 2) %>%
  separate(bigram, into = c("word1", "word2"), sep = " ") %>%
  filter(
    !word1 %in% all_sw,
    !word2 %in% all_sw,
    str_length(word1) > 1,
    str_length(word2) > 1
  ) %>%
  count(word1, word2, sort = TRUE) %>%
  slice_head(n = 30)

bigram_edges
knitr::kable(bigram_edges)

```

```{r}
# Visualize bigram relationships as a network graph

# Load packages required for network graphs
library(igraph)   # creates network objects
library(ggraph)   # visualizes networks using ggplot2
library(ggplot2)

# Convert the bigram table into a graph structure
# - nodes = words
# - edges = connections between word1 and word2
# - edge weight = frequency of the bigram (n)
bigram_graph <- graph_from_data_frame(bigram_edges)

# Plot the network
ggraph(bigram_graph, layout = "fr") +   # Fruchterman-Reingold layout spreads nodes naturally
  
  # Draw connections between words
  # Edge width reflects how often the bigram appears
  geom_edge_link(aes(edge_width = n), alpha = 0.5) +
  
  # Draw nodes representing individual words
  geom_node_point(size = 4, color = "#2C7FB8") +
  
  # Label each node with the corresponding word
  geom_node_text(aes(label = name), repel = TRUE) +
  
  # Remove axes for a cleaner network visualization
  theme_void() +
  
  # Title for the visualization
  labs(title = "Bigram network of Hespress culture articles")
```

The network confirms what the frequency tables already suggested: bigrams cluster around a small number of institutional and national hubs. The thickest edges — reflecting the most frequent pairs — link words like "محمد، السادس، الملك" (the King) and "وزارة، الثقافة، وزير" (Ministry of Culture), underscoring the centrality of state actors in cultural coverage. Clusters around "المهرجان الدولي" (international festival) and "السينما المغربية" (Moroccan cinema) point to recurring cultural events and domains. The fragmented structure of the network — many small, disconnected clusters rather than one dense web — reflects the diversity of topics covered, each article gravitating around its own specific lexical orbit.

We reinforce this with another type of visualisation.

```{r}
# Visualize bigram relationships as a heatmap
library(ggplot2)
ggplot(bigram_edges,
       aes(x = word1, y = word2, fill = n)) +
  # Draw tiles representing each bigram pair
  # Each tile corresponds to one word1–word2 combination
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "#fee090",
    high = "#d73027"
  ) +
  theme_minimal() +
  labs(
    title = "Bigram Co-occurrence Heatmap",
    x = "First word",
    y = "Second word",
    fill = "Count"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

The heatmap shows that bigram co-occurrence in Hespress culture articles is highly sparse and clustered, with only a limited set of word pairs appearing frequently. Overall, the pattern points to structured, repetitive framing in cultural reporting, with a few dominant bigram relationships driving most observed co-occurrence.

## TF–IDF

Some words appear often within a specific article but rarely elsewhere in the corpus. TF-IDF (Term Frequency–Inverse Document Frequency) is a scoring method that identifies these distinctive words. It rewards terms that are frequent in one article but uncommon across all articles. This helps reveal what makes each article topically unique.

TF–IDF is useful here as a complement to raw word frequencies. While frequency counts tell us what words dominate the corpus overall, TF–IDF reveals the other end of the spectrum, what makes individual articles distinct.

```{r}

library(tidytext)

# Compute TF–IDF scores for words in the article bodies

tfidf <- tokens_body %>%
  
  # Count how many times each word appears in each document
  count(doc_id, word, sort = TRUE) %>%
  
  # Compute TF–IDF:
  # TF (term frequency) = how often a word appears in a document
  # IDF (inverse document frequency) = how rare the word is across documents
  bind_tf_idf(word, doc_id, n) %>%
  
  # Order words by decreasing TF–IDF score
  arrange(desc(tf_idf))

# Keep the 20 words with the highest TF–IDF scores
# These represent the most distinctive or document-specific terms
top_tfidf <- tfidf %>% 
  slice_head(n = 20)

# Display the resulting table
kable(
  top_tfidf,
  col.names = c("Document ID", "Word", "Term Frequency", "TF", "IDF", "TF–IDF"),
  caption = "Top TF–IDF terms (Hespress – Culture articles)"
)

# Visualize the top TF–IDF terms
library(ggplot2)

ggplot(top_tfidf, aes(x = reorder(word, tf_idf), y = tf_idf, fill = tf_idf)) +
  # Create a bar chart where bar height corresponds to TF–IDF score
  geom_col() +
  coord_flip() +
  # Flip coordinates so words appear on the vertical axis
  # This improves readability for text labels
  scale_fill_gradient(low = "#9ecae1", high = "#de2d26") +
  labs(
    title = "Top TF–IDF terms (Hespress – Culture)",
    x = NULL,
    y = "TF–IDF score",
    fill = "TF–IDF"
  ) +
  theme_minimal()


```

The highest-scoring terms here are mostly proper nouns and transliterated names (هاليداي، بوي، هيفنر), each concentrated in a single article. This simply tells us that these specific articles are built around a particular artist or cultural figure, it says nothing about how common this pattern is across the corpus.

## Lexical diversity

Lexical diversity measures how varied the vocabulary of the corpus is. The Type–Token Ratio (TTR) compares the number of unique words (*types*) to the total number of word occurrences (*tokens*). A higher ratio indicates a more diverse vocabulary, while a lower ratio suggests more repetition of the same terms. In the context of the Hespress culture corpus, this metric provides a simple indicator of the linguistic richness and variation present in the articles.

The corpus contains approximately 2.9 million tokens and 223,062 unique word types, yielding a type–token ratio (TTR) of 0.077. This relatively low TTR is expected for a large-scale news corpus, as the denominator (total tokens) increases much faster than the number of new word types. Substantively, this indicates a high degree of lexical repetition, driven by the recurrent use of common vocabulary across articles.

```{r}
library(knitr)
lex_div <- tokens_body %>%
  summarise(
    
    # Total number of word tokens in the corpus
    total_words = n(),
    
    # Number of distinct words (vocabulary size)
    unique_words = n_distinct(word),
    
    # Type–Token Ratio (TTR): measure of lexical diversity
    # proportion of unique words relative to total words
    type_token_ratio = unique_words / total_words
  )

# Display the results as a formatted table with a title
kable(
  lex_div,
  digits = 3,
  caption = "Lexical Diversity of Hespress Culture Articles (Type–Token Ratio)"
)

```

# Hypothesis 1

## Sentiment Analysis

Cultural articles that engage more directly with institutional actors or formal domains are expected to display a more emotionally restrained tone, reflected in lower sentiment intensity, compared to culturally oriented articles that lack institutional framing.

To test this hypothesis, we will need:

1.  **Sentiment score per article**

2.  **Institutional proximity measure per article**

```         
binary (institutional vs not) 
```

3.  **Sentiment intensity** `abs(sentiment_score)`

4.  A comparison: descriptive and regression (optional but good)

Sentiment analysis was implemented in Python using the pre-trained **CAMeL-Lab Arabic BERT model for sentiment analysis** (bert-base-arabic-camelbert-da-sentiment) available on Hugging Face. The base CAMeLBERT-DA model was pre-trained on a large corpus of dialectal Arabic web text, comprising approximately **54 GB of text (5.8 billion words)**. The sentiment analysis model does not introduce additional pre-training data, but instead fine-tunes this base model on manually annotated sentiment datasets (ASTD, ArSAS, SemEval), which include both dialectal and Modern Standard Arabic texts.

While the authors provide detailed information on corpus size and language variant, the exact genre composition of the dialectal Arabic training data is not exhaustively specified.

Article texts were tokenized and processed through this transformer-based model to generate document-level sentiment classifications, leveraging Python’s NLP ecosystem for efficient handling of large-scale text. Once each article received a sentiment score and intensity, the resulting dataset of scores was exported (e.g., as CSV) and re-imported into RStudio. In RStudio, these sentiment scores were merged with the original corpus metadata, enabling further descriptive analysis, statistical modeling, and visualization alongside other text features.

The sentiment model outputs probabilities over discrete sentiment classes (positive, negative, neutral). These probabilities are transformed into a continuous sentiment score defined as the difference between the predicted probability of positive sentiment and negative sentiment. Positive values indicate a positive tone, negative values indicate a negative tone, and values close to zero indicate neutrality or balance. Sentiment intensity is defined as the absolute value of this score and reflects the strength of sentiment regardless of direction.

```{r}
#we load the dataset with only the culture category to use it in python
#code available on repo
write.csv(
  hespress_culture[, c("doc_id", "Body")],
  "hespress_culture_body.csv",
  row.names = FALSE
)

```

### **Building the Institutional alignment proxy**

To measure how institutionally oriented each article is, we construct a proxy variable called "institutional alignment". The idea is straightforward: articles that reference state actors, official bodies, or formal discourse markers are considered institutionally aligned.

We define a dictionary of institutional terms across four categories: state and government actors (e.g., the King, ministries, parliament), justice and security bodies (e.g., courts, police, armed forces), public agencies and official institutions, and formal discourse markers (e.g., official statements, decrees, laws).

For each article, we count how many times any of these terms appear in the title and body combined. An article is then flagged as institutional if it contains at least three such hits, a threshold chosen to capture sustained institutional language rather than incidental mentions.

```{r}
library(dplyr)
library(stringr)
# Define a dictionary of institutional and governmental terms

inst_terms <- c(
  # state / government core
  "الملك","القصر","الديوان الملكي","أمير المؤمنين",
  "الحكومة","رئيس الحكومة","الوزير","وزارة","البرلمان","مجلس النواب","مجلس المستشارين",
  "الأمانة العامة","الولاية","العمالة","العامل","الوالي","الجماعة","المجلس الجماعي",
  "الجهة","المجلس الجهوي",
  
  # justice / security
  "المحكمة","النيابة العامة","القضاء","المجلس الأعلى للسلطة القضائية",
  "الشرطة","الأمن","الدرك","القوات المسلحة","الجيش","القيادة","المخابرات",

  # public agencies / official bodies (common)
  "المندوبية","المفوضية","الهيئة","المجلس الوطني","المجلس الأعلى","مؤسسة عمومية","الوكالة","المرصد","اللجنة","اللجان",
  "وزارة الثقافة","مؤسسة ","وزارة التربية","وزارة التعليم","وزارة الصحة","وزارة الداخلية",

  # official discourse markers
  "بلاغ","بيان","تصريح رسمي","مصدر رسمي","وفق\\s+بلاغ","حسب\\s+بلاغ","أفاد\\s+بلاغ",
  "قرار","مرسوم","منشور","قانون","مشروع\\s+قانون","مذكرة"
)

```

This step converts the list of institutional terms into a single regular expression pattern, allowing the corpus to be searched efficiently for any occurrence of these terms.

```{r}
inst_regex <- str_c(inst_terms, collapse = "|")
# Combine all institutional terms into a single regex pattern to match any of them in the text
```

This step operationalizes institutional discourse detection in the corpus. By counting how often terms from the institutional dictionary appear in each article, the code produces a simple indicator of institutional presence.

```{r}
# Build a single text field by concatenating Title + Body (handling missing values),
# then count how many institutional keywords appear (inst_hits),
# and flag the article as "institutional" if it contains at least 2 hits.
hespress_culture <- hespress_culture %>%
  mutate(
    Body = str_c(coalesce(Title, ""), " ", coalesce(Body, "")),
    inst_hits = str_count(Body, inst_regex),
    institutional_article = inst_hits >= 3
  )

```

Articles are classified as institutional if they contain **at least three institutional keyword occurrences**, which is intended to capture sustained rather than incidental institutional language. Under this definition, 2187 articles are flagged as institutional, while 5622 articles contain zero or one institutional keyword and are classified as non-institutional.

Institutional keyword counts are generally low across the corpus. The median number of hits is one, and 75% of articles contain three or fewer institutional keywords, indicating that most cultural articles either do not reference institutions or do so only minimally. The mean count (≈2.23) exceeds the median due to a strongly right-skewed distribution driven by a small number of articles with many institutional references, with extreme cases reaching up to 69 hits. Overall, this pattern suggests that sustained institutional framing is present in a substantial minority of articles but is not the dominant mode of discourse in cultural reporting.

```{r}
# Check how many articles contain at least one institutional keyword,
# and summarize the distribution of keyword counts per article.
library(knitr)

table(hespress_culture$institutional_article) %>%
  as.data.frame() %>%
  kable(
    col.names = c("Institutional article", "Number of articles"),
    caption = "Distribution of institutional vs non-institutional articles"
  )

library(dplyr)
library(knitr)
library(tibble)
# Summarize how frequently institutional keywords appear

sum_inst <- summary(hespress_culture$inst_hits)

# Convert the summary output into a tidy data frame
# so it can be displayed in a formatted table.

data.frame(Value = as.numeric(sum_inst)) %>%
  mutate(Statistic = names(sum_inst), .before = 1) %>%
  kable(
    col.names = c("Statistic", "Value"),
    caption = "Summary statistics of institutional keyword counts per article"
  )


```

Now, let's visualise the proxy.

The following histogram shows how institutional keyword counts are distributed across articles.

```{r}
hist(
  hespress_culture$inst_hits,
  main = "Distribution of Institutional Keyword Counts",
  xlab = "Number of institutional keyword hits",
  ylab = "Number of articles",
  col = "#6baed6",
  breaks = 30,
  xlim = c(0, 20)
)
```

The distribution is strongly right-skewed. Most articles cluster at 0–2 institutional keyword hits, with frequency dropping sharply beyond that. By 10 hits the counts are already negligible. This confirms that institutional language, when present, is typically light, only a small minority of articles engage heavily with institutional actors or discourse.

### **Merging the datasets**

Now, back to reuploading our ready to go sentiment analysis dataset.

```{r}
library(dplyr)
library(readr)

## Import the sentiment analysis output file produced earlier

sentiment_df <- read_csv(
  "hespress_sentiment_transformer.csv",
  show_col_types = FALSE
)
View(sentiment_df)
```

This step combines the sentiment analysis output with the main corpus dataset.

```{r}
#we merge with our existing dataset
hespress_culture <- hespress_culture %>%
  left_join(sentiment_df, by = "doc_id")
```

After merging the sentiment dataset with the article dataset, it is important to verify the integrity of the sentiment variable. The summary statistics provide an overview of the distribution of sentiment scores, while the `NA` count checks whether some articles did not receive a sentiment prediction or failed to merge correctly.

```{r}
# Inspect sentiment score variable and check for missing values

library(knitr)

# Compute descriptive statistics for the sentiment score variable
sum_sent <- summary(hespress_culture$sentiment_score)
# Convert the summary output into a table format for reporting
kable(
  data.frame(
    Statistic = names(sum_sent),
    Value = as.numeric(sum_sent)
  ),
  caption = "Summary statistics of sentiment scores"
)

# Count how many sentiment values are missing (NA)
sum(is.na(hespress_culture$sentiment_score))

```

-   **No missing values**: the NA count is 0, so sentiment scores are available for all articles.

-   The sentiment score spans almost the full theoretical range, from **−0.997 to +0.996**, indicating that the model assigns both strongly negative and strongly positive sentiment to some articles.

-   The **median is exactly 0**, which suggests that the typical cultural article is sentimentally neutral, or that positive and negative sentiment probabilities are closely balanced for most texts.

-   The **mean is slightly positive (≈ 0.04)**, indicating a small overall tilt toward positive sentiment across the corpus, though the magnitude is modest.

-   The interquartile range (from about −0.36 to +0.46) shows that the middle 50% of articles exhibit mild sentiment, with relatively few extreme values.

Overall, this distribution suggests that Hespress culture articles are predominantly neutral in tone, with limited polarization and only a small subset exhibiting strong positive or negative sentiment.

```{r}
hist(
  hespress_culture$sentiment_score,
  breaks = 40,
  main = "Distribution of Sentiment Scores",
  xlab = "Sentiment score",
  col = "lightblue",
  border = "white"
)
legend("topright", 
       legend = "Sentiment score per article", 
       fill = "lightblue")
```

The distribution of sentiment scores is highly concentrated around zero, with a large mass of neutral observations and relatively fewer strongly positive or negative cases. This non-normal, bounded distribution is consistent with the construction of the sentiment measure and reflects the predominantly neutral tone of cultural reporting.

Because the sentiment score is bounded and derived from class probabilities, departures from normality are expected and do not invalidate linear modeling of conditional mean differences.

## **Descriptive analysis**

```{r}
library(dplyr)
library(knitr)
# Compare sentiment between institutional and non-institutional articles

hespress_culture %>%
  group_by(institutional_article) %>%  # Group the dataset by the institutional_article indicator
  summarise(
    mean_sentiment = mean(sentiment_score, na.rm = TRUE),
    sd_sentiment   = sd(sentiment_score, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  kable(
    col.names = c("Institutional article", "Mean sentiment", "SD sentiment", "N"),
    caption = "Mean sentiment by institutional framing"
  )

```

We first split the Hespress culture articles into two groups using the binary institutional proxy (`institutional_article`). Articles coded FALSE are those with 0–1 institutional keyword hits (i.e., institution-light or only incidental mentions). Articles coded TRUE are those with ≥ 3 institutional keyword hits, which we treat as indicating sustained institutional language.

Next, we computed the mean sentiment score, the standard deviation, and the number of observations in each group. The institution-light group (FALSE) contains 5622 articles and has an average sentiment score of 0.074 (SD = 0.549), suggesting a slightly positive tone on average. The institution-heavy group (TRUE) contains 2187 articles and has a lower average sentiment score of −0.035 (SD = 0.524), indicating a more neutral-to-slightly negative tone.

Overall, the descriptive comparison shows that articles with sustained institutional language have **lower average sentiment** than those with minimal institutional references (a raw mean difference of roughly **0.11** points), even before introducing controls such as article length.

```{r}
library(ggplot2)

# Visualize how sentiment scores differ by institutional framing

ggplot(hespress_culture,
       aes(x = institutional_article,
           y = sentiment_score,
           fill = institutional_article)) +
  # Draw a violin plot showing the full distribution of sentiment
  # for each category. The width reflects the density of values.
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  scale_x_discrete(
    labels = c("0–2 institutional hits", "≥ 3 institutional hits")
  ) +
  scale_fill_manual(
    values = c("FALSE" = "#B0C4DE", "TRUE" = "#4682B4")
  ) +
  labs(
    title = "Distribution of Sentiment by Institutional Framing",
    x = "Institutional framing",
    y = "Sentiment score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


```

This figure presents a violin plot with embedded boxplots showing the distribution of sentiment scores for institution-light and institution-heavy articles. Descriptively, sentiment distributions overlap substantially across institutional and non-institutional articles, although institutional articles exhibit a slight downward shift toward neutrality.

Conclusion: Taken together, the initial descriptive analyses provide initial support for the hypothesis. While the sentiment distributions of institution-light and institution-heavy articles overlap substantially, articles containing sustained institutional language exhibit **slightly lower average sentiment scores** and a modest shift toward neutrality. The absence of a pronounced visual separation suggests that any association between institutional framing and sentiment is small in magnitude. Accordingly, regression models are required to assess whether this relationship persists once controlling for article length and potentially other confounding factors.

## **Testing H1**

### T-tests

To examine whether institutional framing is associated with differences in sentiment, we proceed in two steps. We first use a Welch two-sample t-test to compare mean sentiment scores between institutional and non-institutional articles. The Welch variant is preferred over the standard t-test because it does not assume equal variances across groups, which is appropriate here given the skewed distribution of institutional keyword counts.

The t-test gives us a clean, direct answer to the core question: do the two groups differ in sentiment on average? However, because institutional framing is correlated with article length (institutional articles tend to be longer) a simple group comparison may confound the two effects. We therefore follow up with a linear regression model that includes article length as a control variable, allowing us to isolate the effect of institutional framing on sentiment independently of text length.

A Welch two-sample t-test indicates a statistically significant difference in mean sentiment between institution-light and institution-heavy culture articles (t = 8.20, df = 4157, p \< 3.29×10⁻¹⁶). Institution-light articles (fewer than 3 institutional hits) have a higher mean sentiment (M = 0.074) than institution-heavy articles (≥3 hits; M = −0.036). The estimated mean difference (M_FALSE − M_TRUE) is 0.110, with a 95% confidence interval of \[0.084, 0.136\], indicating a small but reliably positive difference in average sentiment. Because this comparison is bivariate, subsequent regression models are used to assess whether the association persists after controlling for article length.

```{r}
t.test(
  sentiment_score ~ institutional_article,
  data = hespress_culture
)
```

### **Regression**

We estimate a linear regression model in which sentiment is regressed on institutional framing, controlling for article length.

Article length is included as a control variable because longer texts tend to exhibit lower sentiment scores, both mechanically and substantively. Mechanically, longer articles contain a higher proportion of neutral and informational language, which dilutes emotionally valenced content and pulls sentiment scores toward zero. Substantively, institutional density is itself correlated with article length, as institutionally framed articles typically provide more background and procedural detail. Controlling for article length therefore allows the effect of institutional density on sentiment to be interpreted independently of differences in text length.

```{r}
#creating the article length variable
library(stringr)
library(dplyr)

hespress_culture <- hespress_culture %>%
  mutate(
    Body = str_c(coalesce(Title, ""), " ", coalesce(Body, "")),
    n_words = str_count(Body, "\\S+")
  )

```

```{r}
#runninng the linear regression

model1 <- lm(
  sentiment_score ~ institutional_article + n_words,
  data = hespress_culture
)
# Load packages for model output formatting and table styling
library(broom)
library(kableExtra)
tidy(model1, conf.int = TRUE) %>%
  mutate(
    term = recode(term,
      "(Intercept)" = "Intercept",
      "institutional_articleTRUE" = "Institutional framing (≥3 hits)",
      "n_words" = "Article length (words)"
    ),
    p.value = format.pval(p.value, digits = 3, eps = 0.001)
  ) %>%
  kable(digits = 4,
        caption = "OLS Regression: Sentiment Score on Institutional Framing and Article Length",
        col.names = c("", "Estimate", "Std. Error", "t", "p-value", "CI lower", "CI upper")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))



# Display model fit statistics
glance(model1) %>%
  select(r.squared, adj.r.squared, nobs) %>%
  mutate(across(c(r.squared, adj.r.squared), ~round(., 4))) %>%
  kable(
    col.names = c("R²", "Adjusted R²", "N"),
    caption = "Model fit statistics"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

We begin by interpreting the intercept. The intercept represents the expected sentiment score for an article that contains no institutional framing (`institutional_article = FALSE`) and has zero words (`n_words = 0`). Substantively, the "zero words" condition is not meaningful, so the intercept should not be interpreted literally. Instead, it serves as a baseline reference point for interpreting the other coefficients.

We then turn to the main coefficient of interest, institutional framing. Holding article length constant, articles classified as institutional exhibit sentiment scores that are on average 0.064 points lower than non-institutional articles (β = −0.064, SE = 0.014, p \< .001). Given that sentiment scores range from −1 to +1, this coefficient indicates a meaningful shift toward more neutral or negative tone. The result is statistically significant at conventional levels, suggesting that the association between institutional framing and reduced sentiment is unlikely to be driven by random variation.

Next, we interpret the control variable, article length. Article length is negatively associated with sentiment (β = −0.0005 per word, SE = 0.00002, p \< .001), indicating that longer articles tend to display lower sentiment scores. While the per-word effect is small, it accumulates over longer texts and is consistent with the idea that longer articles incorporate more informational and neutral language, thereby dampening overall sentiment intensity.

Finally, we consider overall model fit. The model explains approximately 6.2% of the variance in sentiment scores (R² = 0.062; adjusted R² = 0.061). Although modest, this level of explanatory power is typical in text-as-data research, where sentiment is influenced by many unobserved linguistic and contextual factors.

### **Assumption checks**

#### **Normality of residuals**

**Histogram of residuals**

```{r}
# Plot a histogram of the residuals from model1
# Residuals represent the difference between observed values
# and the values predicted by the regression model.
hist(
  resid(model1),
  breaks = 40,
  main = "Histogram of Regression Residuals",
  xlab = "Residuals",
  col = "pink",
  border = "white"
)

```

We assess the normality of the regression residuals using a histogram. The distribution of residuals is centered around zero but departs from perfect normality. In particular, the residuals exhibit some irregularities and mild multimodality, as well as deviations in the tails. These features indicate that the residual distribution is not strictly Gaussian. However, there is no evidence of extreme skewness or unusually heavy tails, and the residuals remain bounded within a relatively narrow range. Given the large sample size and the bounded nature of the sentiment outcome, such deviations from normality are not unexpected. Moreover, the validity of coefficient estimates in linear regression does not rely on strict normality of residuals, and inference remains reliable in large samples

**Q–Q plot**

```{r}
# Q-Q plot to assess whether model residuals follow a normal distribution
# Points lying on the red line indicate perfect normality
qqnorm(resid(model1))
qqline(resid(model1), col = "red")

```

We further assess residual normality using a normal Q–Q plot. The plot shows that residuals closely follow the theoretical normal line in the central portion of the distribution, indicating approximate normality for the majority of observations. However, deviations are observed in the lower and upper tails, where residuals diverge from the reference line. This pattern suggests departures from strict normality, particularly in the extremes. Such deviations are consistent with the bounded nature of sentiment scores and the discrete structure introduced by a binary predictor.

Given the large sample size, approximate normality in the center of the distribution is sufficient for reliable inference, and these tail deviations are unlikely to meaningfully affect coefficient estimates or hypothesis tests.

#### **Homoskedasticity**

Is the variance of residuals similar across the two groups? Do institutional and non-institutional articles show similar dispersion around their group mean?

```{r}
# Load package for regression diagnostic tests
library(lmtest)
# Perform the Breusch–Pagan test for heteroskedasticity
# This test checks whether the variance of the residuals is constant.
# H0 (null hypothesis): residuals have constant variance (homoskedasticity)
# H1 (alternative): residual variance depends on the predictors (heteroskedasticity)
bptest(model1)

```

We first assess the assumption of homoskedasticity using the studentized Breusch–Pagan test. The test evaluates whether the variance of the regression residuals is constant across observations or instead depends on the model’s predictors. The null hypothesis of homoskedasticity is rejected (**BP = 70.937, df = 2, p \< .001**), indicating that the variance of the residuals is not constant. This result suggests that institutional and non-institutional articles, as well as articles of different lengths, do not exhibit identical dispersion around their predicted sentiment scores. While this finding does not affect the unbiasedness of coefficient estimates, it implies that conventional OLS standard errors may be unreliable. We therefore complement the formal test with a visual inspection of residuals to better characterize the nature of this heteroskedasticity.

**Residuals vs fitted values plot**

```{r}
# Plot model residuals against fitted (predicted) values.
plot(
  fitted(model1),
  resid(model1),
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals vs Fitted Values"
)
# Add a horizontal reference line at zero.
# This helps visually check whether residuals are centered around zero.
abline(h = 0, col = "red")

```

\
This residuals-versus-fitted values plot displays the residuals of the regression against the model’s fitted values.

Several features are immediately apparent. First, the residuals are centered around zero across the range of fitted values, indicating that the model does not systematically over- or under-predict sentiment at different levels of the fitted outcome. However, the vertical dispersion of residuals is **not constant**. In particular, the spread of residuals increases as fitted values move toward one end of the distribution, producing a visible fan- or wedge-shaped pattern. This indicates that the variance of the residuals depends on the level of the fitted values rather than remaining constant.

This pattern provides visual evidence of **heteroskedasticity, consistent with the earlier Breusch–Pagan test results.**

While heteroskedasticity does not bias coefficient estimates in linear regression, it can affect standard errors. Accordingly, heteroskedasticity-robust standard errors are used for statistical inference.

```{r}
library(sandwich)
library(lmtest)

# Re-estimate coefficient tests using robust standard errors
coeftest(model1, vcov = vcovHC(model1, type = "HC1"))

```

Importantly, the estimated coefficients remain unchanged, and the substantive conclusions of the model are unaffected: institutional framing continues to be negatively and statistically significantly associated with sentiment scores, even after controlling for article length. This indicates that the observed relationship is not driven by unequal error variance and is robust to violations of the homoskedasticity assumption.

#### **Linearity**

Because the main predictor is binary, linearity is **mostly trivial**. We can still check it for the continuous control variable.

```{r}
# Keep only rows with complete data for the variables used in model1
keep <- complete.cases(hespress_culture[, c("sentiment_score", "institutional_article", "n_words")])

x <- hespress_culture$n_words[keep]
y <- resid(model1)[keep]   # aligned to the same rows

plot(
  x, y,
  xlab = "Article length (words)",
  ylab = "Residuals",
  main = "Residuals vs Article Length"
)
abline(h = 0, col = "red")

```

Residuals were plotted against the number of words per article. The residuals remain centered around zero across the full range of article lengths, with no clear systematic curvature or trend. This suggests that the relationship between article length and sentiment score is adequately captured by a linear specification. While some increase in dispersion is visible for longer articles, this pattern relates to heteroskedasticity rather than nonlinearity and does not indicate a violation of the linearity assumption.

#### **Independence**

The independence assumption requires that each observation provides unique information and that residuals are not systematically related across observations. In this study, each unit of analysis corresponds to a distinct news article, and articles are treated as independent observations. There is no panel structure, repeated measurement of the same article, or clustering by author or publication date explicitly modeled. As such, no formal statistical test is applied for this assumption, and independence is assumed by design of the data.

#### Conclusion

The inferential analyses provide support for the hypothesis that institutional framing is associated with more neutral or negative sentiment. Both the difference-in-means test and the linear regression results indicate that articles classified as institutional exhibit significantly lower sentiment scores than non-institutional articles. Importantly, this association remains statistically significant after controlling for article length and when inference is based on heteroskedasticity-robust standard errors. However, the overall explanatory power of the model is limited, suggesting that institutional framing accounts for only a small share of the variation in sentiment. Accordingly, the results should be interpreted as evidence of a systematic association rather than an exhaustive explanation of sentiment patterns.

Taken together, these results lend empirical support to the hypothesized relationship between institutional alignment and sentiment tone.

# **Discussion**

This study provides evidence consistent with the proposed hypothesis that institutional framing is associated with more neutral sentiment in cultural news articles in the Hespress dataset. While the observed association is statistically robust across specifications, the analysis is subject to several limitations.

First, the institutional framing proxy is binary and relies on a dictionary-based measure that captures the presence of institutional language but does not distinguish between different institutional actors, rhetorical contexts, or degrees of intensity. In particular, the measure does not account for the relative weight, source, or function of institutional references within an article. Future research could develop more nuanced indicators by differentiating between types of institutions (e.g., governmental, judicial, cultural) or by weighting institutional language according to frequency or contextual prominence in the text.

Second, sentiment scores are derived from a pre-trained language model, which may understate evaluative tone in formal or informational writing. Subsequent studies could address this limitation by comparing multiple sentiment models, incorporating domain-specific fine-tuning, or complementing automated sentiment scores with human-coded validation samples.

Third, the modest explanatory power of the models cautions against strong causal interpretations and suggests that sentiment is influenced by additional linguistic and contextual factors not captured in the present analysis. These may include narrative structure, topical focus, or stylistic conventions, which warrant further investigation.

Finally, the current analysis treats articles as independent units and does not account for potential clustering by publication period or author. Extending the research design to include hierarchical or time-aware models would allow future work to examine whether the association between institutional framing and sentiment varies across contexts. Together, these extensions would help clarify the mechanisms through which institutional language shapes tone and improve the interpretability of sentiment-based analyses in media research.
